Comparison of large-scale citizen science data and long-term study data for phenology modeling

### 2018-09-12 manuscript/Reviewer_Reply.md

**l. 127: phenophase (singular)**

Fixed.

**ll. 210-211. I don't understand this. Why do some seemingly strong relationships (e.g., "naive" panel of Fig. 2; also some panels of Fig. 3) yield negative R-squared values? This probably just reflects my ignorance... but it's likely that if I'm mystified, other readers will be, too! Please explain more clearly how the <0 R-squareds arise.**

Maybe change the Fig 2. caption to "..can be negative if the relationship between the two parameter sets is substantialy worse than a 1:1 relationship"

@@

**This is a very practical study and the evaluation of the approach is clear and pretty straightforward to follow. The authors might improve readability a bit by either adding subheads into the Methods & Results or repeatedly using the same terms to refer to the various steps (eg, cross-validation).**

After some discussion, we decided not to add subheadings.

**The most critical thing that I see that’s lacking is that there is no consideration of how decisions made regarding which NPN data to include in the analysis might be impacting model performance. The authors allowed observations with up to 30d between the last reported “no” and the first reported “yes” and chose the midpoint. This could still lead to up to 15d error in the assumed date of “yes” and the true date of “yes”; this variance may in many cases be larger than the signal in the model. Though actually addressing this sensitivity is probably beyond the scope of the analysis, the potential and implications for this error should be discussed explicitly in the paper.**

We tested the sensitivity of this cutoff time by repeating the analysis with a 15 day cutoff. We found no significant differences throughout the analysis. Three species (flowers and budburst for Vaccinium parvifolium and  budburst for Acer circinatum) were dropped from the 15 day analysis due to inadequate sample size. The overall sample size for the 15 day analysis was decreased by 10% (619 observations) and by 1-35% for individual species. We have included notes on this sensitivity test in the methods and results as well as additional supplementary material.

Potential discussion on the technical points:

Gerst et al. 2016 looked at the sensitivity of different thresholds in the USA-NPN dataset and found that differences in site level means among more restrictive datasets was normally distributed. We feel having a shorter cutoff period here would result equally in less overestimates as well underestimates, keeping the optimization surface of the model essentially the same. This is supported by the parameter estimates being nearly identical for the majority of cases (see new images S1-S3).

@@

Shawn's note:
After talking with Ethan I'll redo most of the analysis with a shorter cutoff date and see how things compare.
also report the distribution of "days between" to see if they're very far off between 30day cutoff and a shorter cutoff.

**The authors lay out two goals for this effort – to determine “inferences about biological processes driving phenology” and to develop predictive phenology models. It may be just me…using the term “inference” in this way confused me.**

**Regarding the predictions that can result from these different models…this is of strong interest to many stakeholder realms. It would be fantastic to see a bit more attention given in the discussion to recommendations for implementation. The authors talk about the tradeoffs between more complex formulations and simpler models and – generally – reasons these differences may come about. It would be great to also see a bit of more practical discussion along the lines of, “if you can tolerate error of 7-14d, then model X may meet your needs, and it’s fine to build the models with either NPN or LTER data. However, if your needs are Y, then you’d be better off doing blah, blah…”… would this be possible?**

We have added an additional note describing the best models overall from our analysis and pointed readers to the supplemental images for specific details of model performance. We have also updated supplemental images S1 and S2 to better show model performance in the cross-dataset comparison, and also added an additional table (Table S2) which gives the overall model performance across all species and phenophases.

@@

**I think it’s pretty critical to provide your sample sizes for each of the species/phenophases under evaluation for each of the datasets… when I read that you held out 20% of observations for evaluation, I was left wondering, was this 5 points or 50 or 500? You could include this info in Table 1**

We have added details of the sample sizes to the supplement table S1.

**What is the sampling frequency for the LTER sites? This is quite relevant because you chose to use 30d as your cut-off for NPN data - you’ll allow for up to 30d between the last reported “no” and the first reported “yes”. How does this compare to what’s possible in the LTER datasets? It could definitely have an impact on the sensitivity of your results.**

Three of the four LTER sites (Harard, Hubbard Brook, & H.J. Andrews) have very short sampling intervals (3-7 days). The Jornada LTER has a sample interval of 30 days. We have added these details to the supplementary methods.
With the exception of the Jornada we feel the sampling interval of the LTER sites is very precise, and is one of the strengths of long term data that we point out throughout the manuscript. Given that, the issue of potentially 30 (or 15) days between sampling versus 7 is one of several points relating to volunteer based sampling we discuss in the manuscript.

We note that only 2 of the 38 comparisons are from the Jornada datasets, so it's larger sampling frequency does not affect the overall analysis done in this paper.

@@

This referred to the RMSE values derived from LTER Model predictions for held out LTER data. We rewrote this paragraph slightly and changed the labels on Figure 4 to make this analysis more clear.

### Results

**L231-237: is it possible to quantify these comparisons in any way?**

We have added a t-test to show the difference from 0 in these distributions.

@@

**L244: are you referring to the temporal depth and/or frequency of sampling when you say “intense”?**

We are referring to both of these as LTER sites typically have a long history of frequent and consistent sampling compared to USA-NPN sites.

**263: “limitations in the sampling of the NPN dataset” – what sort of limitations? Small sample size, frequency of sampling, or something else?**

The frequency of sampling, in a given season and also between seasons, is the primary limit we consider here and we've added this clarification in the text.

@@

**While I do think that this manuscript should be ultimately acceptable for publication at Ecology, there are a few major/minor comments and suggestions for which I would like the authors to consider.**

Major Comments and Suggestions:

**First, for NPN species with larger pools/spatial extents of observations, I wonder if would be useful to calibrate the models using subsets somehow stratified by climate and/or latitude. For example, train the models using only warm region (or southern) red maple and test on cold region maples (or vice versa). Perhaps this could further address the issue of model identifiability (as discussed in Lines  259-277). While the oft-mentioned Basler (2016) study certainly had nice spatiotemporal coverage, it did not have nearly the latitudinal/climate variability that NPN affords.**

We discussed this and ultimately decided that it is beyond the scope of the current paper. Although we agree that this approach has the potential to improve the paper, we are doubtful if it would address the issue of identifiability. The Harvard and Hubbard Brook LTER share 3 species in common and are relatively close (~150km), yet their parameter estimates varied widely for most models (see images S4-S5 in the original and revised text). 
We also feel that it would require too much space to adequately investigate and discuss, and that the current set of analyses sufficiently accomplish the goals we set out in the introduction.

@@

**Second, given the length of time series of the LTER datasets, I think it would be useful to further examine the performance of models based on other criteria such as the amount of interannual variability explained (correlation or R-square) or bias during anomalously warm or cold years. Perhaps the latter might be outside of the overall realm of the paper, but I think it would be informative to know which models capture this type of information which matters in a global change context.**

See the note to Reviewer 1 about general model recommendations. We have updated the Supplemental images S1 and S2 to included Pearson's correlation in addition to the RMSE for all dataset inter comparisons.

@@

### Minor Comments and Suggestions:

**Line 281: Please verify what is T* (I assume base temperature)**

This is correct. We have changed T* to T_base throughout as it is more descriptive.

### 2018-09-12 manuscript/main.tex
final revisions edits from shawn before sending to co-auths

In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology and our ability to predict future phenology events. Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the USA-NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network. We compare the resulting models based on both inference about models and parameters and predictions for unobserved events. We then use this comparison to assess the best methods for both local- and large-scale phenology modeling and to point the way forward for integrating large-scale and local-scale data to determine the best possible models across scales.

@@

The USA National Phenology Network protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the USA-NPN datasets are located across the U.S. and generally clustered around populated areas (Fig. 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig. 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

We downloaded all USA-NPN observations from 2009, when collections began, to 2016 for the following phenophases: Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers \citep{npndata}. The first three phenophases apply to the 'leaf out' phase for deciduous broadleafs, evergreen conifers, and pines, respectively. The 'Open Flowers' phenophase refers to fully-open flowers and applies to all angiosperms. Hereafter, we will refer to these as either 'Flowers' for the Open Flower phenophase, or 'Budburst' for all other phenophases. We subset the USA-NPN observations similar to methods outlined in \cite{crimmins2017}. First, 'yes' observations for individual plants were kept only if they were preceded by a 'no' observation within 30 days. Observations for 'Budburst' that were past day of year (DOY) 172, and for 'Flowers' that were past DOY 213 were dropped to minimize any influence from outliers. We inferred the observed DOY of each phenophase as the midpoint between each 'yes' observation and the preceding 'no' observation. Finally, only species that had greater than 30 total observations were kept. \cite{crimmins2017} only kept observations that were preceded by a 'no' within 15 days, and also grouped multiple individuals at single sites to a single observation. We used 30 days to allow for a greater number of species to be compared. We chose not to group multiple individuals at a single site to better incorporate intra-site variability.

In the LTER datasets observation metrics varied widely due to different protocols. To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1).

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (USA-NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature and are calculated differently for each model (Table 2). The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T$ \citep{reaumur1735, wang1960, hunter1992}. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T$ = 0$^{\circ}$C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 (number of chill days - $NCD$). The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.

@@

We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations. Using the RMSE values from held out observations, we  compared the  performance of LTER and USA-NPN derived models on different data types in two different ways. First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and USA-NPN derived models solely with LTER observations. Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely USA-NPN data. These differences were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the USA-NPN derived model performed better. We used a t-test to test the difference from zero in these values. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available. Absolute RMSE values are provided in the supplement (Fig. S1-S3).

@@

The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$). The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig. 2). The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Fig. 2). The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}. While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig. S6-S7).

@@

When comparing the prediction accuracy on held-out data, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all p-values < 0.001, Fig. 4). The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations. Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall. The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among USA-NPN derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig. S1 \& S2).

@@

Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T^*$=0 and $t_{1}$=5, $F$=5, $T^*$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig. S6-S7). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S4-S5). These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates which were not biologically realistic \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and USA-NPN datasets. In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions. Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Fig. S1 \& S2). This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data. Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}. In combination with our results showing observed differences in parameter estimates between LTER sites (Fig. S4-S5), this suggests that variation in phenological requirements across the range is likely important. However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses. Specifically, correspondence between parameter estimates (Fig. 2), estimates of phenological events (Fig. 3), and out-of-sample error rates (Fig. 4) for the \texit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively. This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the USA-NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the \textit{M1} model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by having a 40-year time-series across a large region. \cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research. Large-scale phenology datasets, like USA-NPN, will be key in addressing this and other phenological research needs.

@@

\newpage
\begin{adjustbox}{angle=90}
\small
%%{\def\arraystretch{2}\tabcolsep=5pt
    \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
    \hline
    Name & DOY Estimator & Forcing Equations & Total\newline Parameters & Reference \\ \hline
    Naive & \( \overline{DOY} \) & - & 1 & - \\
    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = max(T_{i}, 0)$ & 1 & \citep{reaumur1735, wang1960, hunter1992} \\
    Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T^{*}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T^{*}, 5) $  & 4 & \citep{blumel2012} \\
    Alternating & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
    MSB & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
    \hline
    \end{tabular}
\end{adjustbox} \newline 
\textbf{Table 2}: Phenology models used in the analysis

>>>>

In order to accurately model and forecast phenology, it is important to understand how the strengths and weaknesses of intensive local studies and large-scale citizen science projects influence both our inferences about biological processes driving phenology (eg. warming requirements for a specific plant) and our ability to predict future phenology events (eg. forecasting when flowering or leaf out occurs). Here, we fit a suite of plant phenology models for the budburst and first-flowering phenophases of 24 plant species to data from both the USA-NPN and a set of intensive long-term studies from the Long Term Ecological Research (LTER) network. We compare the resulting models based on both inference about models and parameters and predictions for unobserved events. We then use this comparison to assess the best methods for both local- and large-scale phenology modeling and to point the way forward for integrating large-scale and local-scale data to determine the best possible models across scales.

@@

The USA National Phenology Network (USA-NPN) protocol uses status-based monitoring, where via a phone app or web based interface observers answer 'yes,' 'no,' or 'unsure' when asked if an individual plant has a specific phenophase present \citep{denny2014}. Phenophases refer to specific phases in the annual cycle of a plant, such as the presence of emerging leaves, flowers, fruit, or senescing leaves. Sites in the USA-NPN datasets are located across the U.S. and generally clustered around populated areas (Fig. 1). To represent long-term, intensive phenology studies we used four datasets from North America representing three major ecosystem types (Table 1, Fig. 1). All four long-term studies are located in the U.S. and are part of the Long Term Ecological Research network (LTER). The Harvard Forest and Hubbard Brook Long Term Experimental Forest are located in the northeastern U.S. and are dominated by deciduous broadleaf species. The H.J. Andrews Experimental Forest is a coniferous forest in the coastal range of the western U.S. The Jornada Experimental Range is in the Chihuahua desert of the southwestern U.S.

We downloaded all USA-NPN observations from 2009, when collections began, to 2016 for the following phenophases: Breaking Leaf Buds, Breaking Needle Buds, Emerging Needles, and Open Flowers \citep{npndata}. The first three phenophases apply to the 'leaf out' phase for deciduous broadleafs, evergreen conifers, and pines, respectively. The 'Open Flowers' phenophase refers to fully-open flowers and applies to all angiosperms. Hereafter, we will refer to these as either 'Flowers' for the Open Flower phenophase, or 'Budburst' for all other phenophases. We subset the USA-NPN observations similar to methods outlined in \cite{crimmins2017}. First, 'yes' observations for individual plants were kept only if they were preceded by a 'no' observation within 30 days. Observations for 'Budburst' that were past day of year (DOY) 172, and for 'Flowers' that were past DOY 213 were dropped to minimize any influence from outliers. We inferred the observed DOY of each phenophase as the midpoint between each 'yes' observation and the preceding 'no' observation. Finally, only species that had greater than 30 total observations were kept. \cite{crimmins2017} only kept observations that were preceded by a 'no' within 15 days, and also grouped multiple individuals at single sites to a single observation. We used 30 days to allow for a greater number of species to be compared. We tested the sensitivity of this choice by also performing the analysis using a 15 day cutoff. We chose not to group multiple individuals at a single site to better incorporate intra-site variability.

In the LTER datasets observation metrics varied widely due to different protocols. To match the USA-NPN data we converted all metrics to binary 'yes' and 'no' observations for each phenophase (see supplementary methods). As with the USA-NPN data, we inferred the date for each phenophase as the midpoint between the first 'yes' observation and most recent 'no' observation, and only kept species and phenophase combinations which had at least 30 total observations. After data processing there were 38 species and phenophase combinations (with 24 unique species) common to both the USA-NPN and LTER datasets to use in the analysis (Table 1 \& S1). Using a 15 day cutoff in the USA-NPN dataset resulted in 35 unique combinations with 23 species .

@@

It is common to fit multiple plant phenology models to find the one that best represents a specific species and phenophase \citep{chuine2013}. For each of the 38 species and phenophase combinations in the five datasets (USA-NPN and four LTER datasets), we fit eight phenology models (Table 2). The \textit{Naive model} uses the mean DOY from prior observations as the estimated DOY. The \textit{Linear model} uses a regression with the mean spring (Jan. 1 - March 31) temperature as the independent variable and DOY as the response variable. For the six remaining models, the general form is based on the idea that a phenological event will occur once sufficient thermal forcing units, $F^{*}$, accumulate from a particular start day of the year ($t_{1}$). Forcing units are a transformation of the daily mean temperature and are calculated differently for each model (Table 2). The start day can either be estimated or fixed. For the \textit{Growing Degree Day (GDD)} model, forcing units are the total degrees above the threshold $T_{base}$ \citep{reaumur1735, wang1960, hunter1992}. The \textit{Fixed GDD model} uses the same form but has fixed values for start day ($t_{1}$ = Jan 1) and temperature threshold ($T_{base}$ = 0$^{\circ}$C). The \textit{Alternating model} has a variable number of required forcing units defined as a function of the total number of days below 0$^{\circ}$C since Jan. 1 (number of chill days - $NCD$). The \textit{Uniforc model} is like the GDD model but with the forcing units transformed via a sigmoid function \citep{chuine2000}.

@@

We also directly evaluated model performance using the four combinations of models and observed data: A) LTER-derived models predicting LTER observations, B) USA-NPN derived models predicting LTER observations, C) LTER-derived models predicting USA-NPN observations, and D) USA-NPN derived models predicting USA-NPN observations. Using the RMSE values from held out observations, we  compared the  performance of LTER and USA-NPN derived models on different data types in two different ways. First, we focused on local-scale prediction by calculating the difference in the RMSE of LTER and USA-NPN derived models solely with LTER observations. Secondly, to focus on large-scale prediction we calculated the difference in RMSE using solely USA-NPN data. These differences were calculated for each of the model types and 38 species-phenophase combinations. Negative values indicate that LTER-derived models perform better, while positive values indicate that the USA-NPN derived model performed better. We used a t-test to test the difference from zero in these values. In the three cases where the same species and phenophase combination occurred in two LTER sites, we made the LTER-LTER comparison within each site, not across sites, to focus on local scale prediction when LTER data are available. Absolute RMSE values are provided in the supplement (Fig. S5-S7).

@@

Throughout the analysis no significant difference was found between a 30 day or 15 day threshold between the first 'yes' and most recent 'no' observation in the USA-NPN dataset. Results presented here reflect the 30 day cutoff, see the supplemtary figures S2-S4 for matching figures using a 15 day cutoff. The best matches between parameter estimates based on USA-NPN and LTER data were the \textit{Fixed GDD model} ($R^2$ = 0.49) and the \textit{Linear model} ($R^2$ = 0.39 for $\beta_{1}$ and -0.05 for $\beta_{2}$). The parameters for all other models had $R^2$ values \textless 0 indicating that the relationship was worse than no relationship between the parameters (but with matching mean parameter values across the two sets of models) (Fig. 2). The \textit{Naive model} showed a distinct late bias in mean DOY estimates for phenological events, likely resulting from the LTER datasets being mostly in the northern United States compared to the site locations of the USA-NPN dataset (Fig. 2). The large outlier for the \textit{Fixed GDD model} is \textit{Larrea tridentata}; this species' flower phenology is largely driven by precipitation, which is not considered in the Fixed GDD model \citep{beatley1974}. While the \textit{Fixed GDD} and \textit{Linear} models showed reasonable correspondence between parameter estimates, all parameters for individual species and phenophase combinations had different distributions between USA-NPN and LTER-derived models (Fig. S10-S11).

@@

When comparing the prediction accuracy on held-out data, USA-NPN derived models made more accurate predictions for held-out USA-NPN observations, and LTER-derived models performed better on held-out LTER observations ( all \textit{p} < 0.001, Fig. 4). The \textit{Naive} and \textit{Linear} models had the largest differences between the two model sets, while the \textit{Fixed GDD model} had relatively similar errors when evaluated on both USA-NPN and LTER held-out observations. Although the \textit{Fixed GDD model} had the highest agreement in accuracy between USA-NPN and LTER-derived models, it was not the best performing model overall. The \textit{GDD} and \textit{Uniforc} models commonly made the best predictions, having the lowest RMSE in 23\% and 40\% of cases among USA-NPN derived models, and 42\% and 32\% of cases among LTER-derived models, respectively (Fig. S5 \& S6).

@@

Despite substantial differences in parameter estimates, LTER and USA-NPN derived models produced similar estimates for phenological events in most cases (Fig. 3). This greater correspondence between predictions than parameters suggests that more complex models may have identifiability issues. For example, two GDD models with parameters of $t_{1}$=1, $F$=10, $T_{base}$=0 and $t_{1}$=5, $F$=5, $T_{base}$=0 produce nearly identical estimates in many scenarios. This possibility is supported by the fact that the highest correlation between parameter estimates is seen in models with only 1 or 2 parameters. In addition, bootstrap results for more complex models suggest a high degree of variability in parameter estimates and potentially multiple local optima in fits to both USA-NPN and LTER data (Fig. S10-S11). Finally, parameter estimates of more complex models are also not consistent among models for the same species when comparing multiple LTER datasets (Fig. S8-S9). These results are consistent with research showing that models failed to estimate the starting day of warming accumulation solely from budbreak time-series, thus producing parameter estimates which were not biologically realistic \citep{chuine2016}. \cite{basler2016} suggests that the key component in phenology models is the thermal forcing, with additional parameters being sensitive to over-fitting. Here, our simplest model, the \textit{Fixed GDD model} which uses only a warming component, had the highest correlation among parameters between LTER and USA-NPN datasets. In combination with the aforementioned studies, our results indicate that caution is warranted in interpreting parameter estimates from complex phenology models regardless of the data source used for fitting the models.

While more complex phenology models appear to have identifiability issues, there is also evidence that they capture useful information, beyond the \textit{Fixed GDD model}, based on their ability to make out-of-sample predictions. Based on the RMSE, the \textit{GDD} and \textit{Uniforc} models produce the best out-of-sample predictions for the majority of species and phenophases at both USA-NPN and LTER datasets (Fig. S5 \& S6). This demonstrates that the more complex models are capturing additional information about phenology, and that some of the differences between datasets result from differences in either the scales or the sampling of the data. Spatial variation in phenological requirements is known to exist in plants \citep{zhang2017}. In combination with our results showing observed differences in parameter estimates between LTER sites (Fig. S8-S9), this suggests that variation in phenological requirements across the range is likely important. However, the models that attempted to address this by incorporating spatial variation did not yield improvements over their base models in our analyses. Specifically, correspondence between parameter estimates (Fig. 2), estimates of phenological events (Fig. 3), and out-of-sample error rates (Fig. 4) for the \texit{MSB} and \textit{M1} models were essentially the same as the \textit{Alternating} and \textit{GDD} models, respectively. This lack of improvement from incorporating spatial variation could be caused either by models not adequately capturing the process driving the spatial variation, the USA-NPN dataset having biases from variation in sampling effort and/or spatial auto-correlation, or some combination of these factors. \cite{basler2016} used the \textit{M1} model to predict budburst for six species across Europe and found it was generally among the best models in terms of RMSE, albeit never by more than a single day. Their result was strengthened by having a 40-year time-series across a large region. \cite{chuine2017} listed the incorporation of spatial variation in warming requirements in models as a primary issue in future phenology research. Large-scale phenology datasets, like USA-NPN, will be key in addressing this and other phenological research needs.

In addition to exploring differences between phenology datasets our analysis provides guidance on which models to use when making predictions at a local scale using models built from large scale data, or vice versa. Among the eight models tested the Uniforc model performed the best overall in the cross dataset comparison in terms of pearson correlation and RMSE (Table S2). One exception was the Fixed GDD model had a slightly higher pearson value when using LTER derived models to make predictions for USA-NPN data. When the scale is the same between predictions and training data (ie. using LTER derived models to make predictions for LTER data) the Uniforc, GDD, and M1 models performed equally well, though the Uniforc had slightly better RMSE (Table S2). The best model for specific species and phenophases varied, but was commonly the Uniforc and GDD models (Figs. S5-S6).

@@

\newpage
\begin{adjustbox}{angle=90}
\small
%%{\def\arraystretch{2}\tabcolsep=5pt
    \begin{tabular}{ | l | c | c | p{1.3cm} | l |}
    \hline
    Name & DOY Estimator & Forcing Equations & Total\newline Parameters & Reference \\ \hline
    Naive & \( \overline{DOY} \) & - & 1 & - \\
    Fixed GDD &$\sum_{t=0}^{DOY}R_{f}(T_{i})\geq F^{*} $  & $R_{f}(T_{i}) = max(T_{i}, 0)$ & 1 & \citep{reaumur1735, wang1960, hunter1992} \\
    Linear & \( DOY = \beta_{1} + \beta_{2}T_{mean} \) & - & 2 & - \\
    GDD & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = max(T_{i} - T_{base}, 0) $  & 3 & \citep{reaumur1735, wang1960, hunter1992} \\
    M1 & $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq (\frac{L_{i}}{24})^{k} F^{*} $ & $ R_{f}(T_{i}) = max(T_{i}-T_{base}, 5) $  & 4 & \citep{blumel2012} \\
    Alternating & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD(t)} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 3 & \citep{cannell1983} \\
    MSB & $\sum_{t=0}^{DOY}R_{f}(T_{i})\geq a + be^{cNCD_{i}} +dT_{mean} $ & $R_{f}(T_{i}) = max(T_{i}-5, 0) $ & 4 & \citep{jeong2013} \\
    Uniforc &  $\sum_{t=t_{1}}^{DOY}R_{f}(T_{i})\geq F^{*} $ & $ R_{f}(T_{i}) = \frac{1}{1 + e^{b(T_{i}-c)}} $ & 4 & \citep{chuine2000} \\
    \hline
    \end{tabular}
\end{adjustbox} \newline 
\textbf{Table 2}: Phenology models used in the analysis
